---
title: 'Lab 1: Comparing Means'
author: "Sristhi Mehra, David Djambazov, and Andi Morey Peterson"
date: "10/17/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, message=FALSE}
library(tidyverse)
A = read.csv("anes_pilot_2018.csv")
```

# The Data

The American National Election Studies (ANES) conducts surveys of voters in the United States.  While its flagship survey occurs every four years at the time of each presidential election, ANES also conducts pilot studies midway between these elections.  We will be using this data to ask five (5) questions about the responsants:

1. Do US voters have more respect for the police or for journalists?

2. Are Republican voters older or younger than Democratic voters?

3. Do a majority of independent voters believe that the federal investigations of Russian election interference are baseless?

4. Was anger or fear more effective at driving increases in voter turnout from 2016 to 2018?

5. (Student Choice)

## General Study Comments (applicable for all questions)

Since all questions draw from the same study sample, it is useful to state comments here that we can refer to for all questions.  For almost any test we run, we will need to determine if the data is i.i.d. and if the respondents to the survey accurately represent the average U.S. voter that we can generalize accordingly. 

_Independence_ - Unless multiple people in the same locale, same family, or same household, for example, are used in the way the survey was conducted, we can safely assume each respondent is independent from another.

_Identically Distributed_ - Once on person has taken the survey, they cannot take the survey again; so the distribution for the next "draw" is changed.  But the change in the population distribution for the next raw is so small, we can safely ignore this effect.

_Generalizability_ - Because this is a modern, paid, opt-in survey, the sample data will only include individuals who have the propensity or financial motivation to complete the survey.  However, the financial impact is small, 21-50 cents for this 30 minute survey (see the ANES User Guide Code Book).  In addition, the survey provided weights in which the survey recommends to use when making inferences to the target population of U.S. adult citizens.

Given these, we can assume the iid assumption is valid and for results we worry about generalizability, we can use the weights to help us on questions in which we are concerned about generalizing to the population. (One concern not mentioned -- the data did not account for people who are ineligable to vote due to a felony).

_Confidence Interval_ - All tests will use a 95% confidence as standard practice.

\newpage

### 1. Do US voters have more respect for the police or for journalists?
#### Conceptualize
The concept is in the question itself.  Do US voters prefer police over journalists?

#### Operationalize
We could have directly asked the survey respondents to pick on or the other, but the survey didn't do this.  Rather is other data we can use to operationalize this question.  First there is the rating question in which respondents were asked to rate different groups such as police and journalists. “How do you rate the police?” and “How do you rate journalists?” These results reside in the variables $ftpolice$ and $ftjournal$.  There is also the question "How concerned are you about violence against people who work in the news media?" However, since there is not a corresponding question about police, we cannot use that in any two-sample test.

In addition to this, the question asks about voters.  The survey provides a few sample questions about the respondent to determine if that respondent was a voter. Variables $turnout18$ and $turnout18ns$ can be used to determine if they were a voter.

#### Exploratory Data Analysis

First, let’s determine if we have enough "US Voters".


```{r}
A$likelyvoter <- 
  ifelse((A$turnout18<=3), 1, 
         ifelse((A$turnout18==5)&&(A$turnout18ns=1),1,0))
paste("Number of Definite and Probable Voters:", sum(A$likelyvoter==1))
```

We have 1842 people who claimed they definitely or probably voted in which we will use to filter the other two variables $ftpolice$ and $ftjournal$.  Now we need to see if the sample distribution of $ftpolice$ and $ftjournal$ are normal enough to use CLT.

```{r, fig.width=8, fig.height=4}
filtered_voter18<-select(A, c("likelyvoter", "ftpolice", "ftjournal"))
filtered_voter18<-filter(filtered_voter18, likelyvoter==1)
hist(filtered_voter18$ftpolice, breaks = 50,
     main = "2018 Voter Rating of the Police",
     xlab = "Feeling Thermometer  0-100")
hist(filtered_voter18$ftjournal, breaks = 50,
     main = "2018 Voter Rating of Journalists",
     xlab = "Feeling Thermometer  0-100")
```

The histograms reveal that on both the police and journalists there are spikes at 100 and are not normal  Because we will be comparing the means of both of these distributions, and because the number of observations are large enough at over 1800 it isn't a concern; we can rely on CLT.  There are 2 appearances of -7 for the second variable, but since there's only 2 it shouldn't affect the data.

#### The Test
Since our sample's variables are ordinal, i.i.d (see general assumptions on page 1) and the data can be paired, we would argue that using a sign test is the most appropriate way test.  Therefore the _Null Hypothesis_ is that the difference in how 2018 US Voters rate police and how they rate journalists is equal to 0.

$$ \mu_{ftpolice} - \mu_{ftjournal} = 0 $$
```{r}
t.test(filtered_voter18$ftpolice, filtered_voter18$ftjournal)
```
Because the p-value is so low at $p<2.2e^{-16}$, this test is highly significant and we can reject the null hypothesis that US Voters rated journalists and police the same. We can compare the means to understand the effect size.  The mean rating for police is 68.491, while the mean rating for journalists is 54.418, so US voters, on average, rate the police 14 "temperature points" higher than journalists.

\newpage

### 2. Are Republican voters older or younger than Democratic voters?
#### Conceptualize
Here we are trying to understand if there is an age component in the differences in political parties.

#### Operationalize
The survey itself asks its respondents many questions surrounding their political identity. We have several variables in which we can consider to use to define political identity. First, there is $pid1d$ / $pid1r$ which asks "Generally speaking, do you usually think of yourself as a Democrat, a Republican, an Independent, or what?".  Each of these reverse the order of which is asked first.  There are follow up questions with these to try and figure out which way the respondent leans or how strongly $pidstr$ and $pidlean$.  There is also $pid7x$ that also asks on a Likert-type scale. Since we are asking just ages of Republican vs Democrat and not HOW far they lean or how STRONGLY they feel that way, we can use the data from $pid1d$/$pid2d$ (jointly, since 50% were asked one way and 50% the other).  We will disregard WHO they voted for, as the person they voted for doesn't represent which party they are affiliated with.  For age, we can simply use $birthyr$ and know that we will only have a year granularity for our data (rather than month and day).

#### Exploratory Data Analysis

First, lets determine if we have enough Democrats and Replublicans.  We will have to awknowlege that using only reponses 1 or 2 for $pid1d$ and $pid2d$ will remove non-responses, Indpendants, and "others" from the analysis.  We will also have to confirm these individuals are voters.

```{r}
Age_and_Party <-select(A, c("pid1d", "pid1r", "likelyvoter", "birthyr"))
Age_and_Party$Party <- 
  ifelse((Age_and_Party$pid1d==2 | Age_and_Party$pid1r == 2), 0,
  ifelse((Age_and_Party$pid1d==1 | Age_and_Party$pid1r == 1), 1, -1)) 
Age_and_Party<-filter(Age_and_Party, Party >= 0)
Age_and_Party<-filter(Age_and_Party, likelyvoter >= 1)
paste("Number of Republicans: ", sum(Age_and_Party$Party==0))
paste("Number of Democrats:   ", sum(Age_and_Party$Party==1))
```

We have 1228 people who claimed they are part of either party, Democrates and Republicans.  Now we need to see if the sample distribution of their ages are normal enough to use CLT.

```{r, fig.width=8, fig.height=4}
hist(Age_and_Party$birthyr, breaks = 50,
     main = "Histogram of Democrats and Republicans Birth Years",
     xlab = "Year")
```

The historgram of ages has a somewhat bimodal distribution however, with 1228 reponses, we can rely on CLT. 

#### The Test
Since our sample's political id variable is metric (0 or 1) and birthyr is metric, i.i.d (see general assumptions on page 1) and the data cannot be paired, we would argue that using an unpaired t-test.  The _Null Hypothsis_ is that the difference ages between Democrats and Republicans is equal to 0.
$$ \mu_{D\_age} - \mu_{R\_age} = 0 $$
```{r}
t.test(Age_and_Party$birthyr~Age_and_Party$Party)
```
Here we have a statistically significant test is the p-value being less than 0.05.  We can reject the null hypthosis that there is an no age difference between Democrats and Republicans voters.  Republican voters are, on average, 3 years older than Democrat voters since they have a mean of a birth year of 1962 and the Democrat voters have a birth year mean of 1965.

\newpage
### 3. Do a majority of independent voters believe that the federal investigations of Russian election interference are baseless?
#### Conceptualize
The concept here is to understand if indpendent voters support the Russian election investigations.

#### Operationalize
As with the second quesiton, we can use $pid1d$ and $pid2d$ to determine if a voter self-identifies as an independant voter.  There are a few ways we can operationalize their belief in the Russian Election Interference. The variable $russia16$ asks - "Do you think the Russian government probably interfered in the 2016 presidentia election to try to help Donald Trump win, or do you think this probably did not happen?" The variable $muellerinv$ asks "Do you approve, disapprove, or neither approve nor disapprove of Robert Mueller’s investigation of Russian interference in the 2016 election?"  Both of these can get to the question of if a voter things the investigation is baseless. The first is a simple "yes" or "no" type question, making for an easier type of analysis.  Let's try this one first...


#### Exploratory Data Analysis

First, lets determine if we have enough  Independant voters.  We will have to awknowlege that using only reponse selection 3 for $pid1d$ and $pid2d$ will remove Democrats, Republicans, and "others" from the analysis.  We will also have to confirm these individuals are voters.

```{r}
Independant_Voters <- select(A, c("pid1d", "pid1r", "likelyvoter", "russia16"))
Independant_Voters <- filter(Independant_Voters, pid1r == 3 | pid1d == 3)
paste("Number of Independants: ", nrow(Independant_Voters))
paste("Number of nonreponse to Russia quesiton", sum(Independant_Voters$russia16<0))
```

We have 767 people who claimed they are Independant AND Voted AND conviently also gave an answer to the Russia question.

Since our sample's variable is metric (1-believe or 2-disbelieve), i.i.d (see general assumptions on page 1) and the data cannot be paired, we would argue that using a one-sample t-test.  The _Null Hypothsis_ is that the 50% of the respondants woudld pick 1 and 50% woudl pick 2, making the mean 1.5.
$$ \mu =1.5 $$
```{r}
t.test(Independant_Voters$russia16, mu=1.5)
r = (sum(Independant_Voters$russia16==1) - sum(Independant_Voters$russia16==2))/(nrow(Independant_Voters))
paste("r value: ", r)
```
Here we have a statistically significant test is the p-value being less than 0.05.  We can reject the null hypthosis that the question was answered 50% one way and not the other.  We can accept the alternative hypothesis that it mu is not 1.5.  More people believe the invetigation was warrented than not but not by much, as the r-value (correlation) is very low at 9.7%.

\newpage
### 3. Was anger or fear more effective at driving increases in voter turnout from 2016 to 2018?
#### Conceptualize
The concept here is to understand what drove increases in voter turnout from 2016 to 2018 in terms of just anger or fear. 

#### Operationalize
We have data on who voted in 2016 in $turnout16$ and who voted in 2018.  Our population that we are curious about are those who did NOT vote (or couldn't remember if they did) in 2016 but who are confident that they did vote in 2018.  The survey also collected data on those voters on how they are generally feeling.  Most emotions were listed, inlcuding anger and fear.  This variable $geangry$ and $geafraid$ are collected on a scale from 1-5, ranging from Not at all [1], A little [2] Somewhat [3] Very [4] and Extremely [5].  We can operationalize those who are angry or scared as signifiying the said Very or Extermely for the answer to this question.


#### Exploratory Data Analysis

First, lets determine if we have enough voters who did not, or were not sure they voted in 2016 but then did in 2018. 

```{r}
Non2016Voters <- select(A, c("likelyvoter", "turnout16", "geangry", "geafraid"))
Non2016Voters <- filter(Non2016Voters, turnout16 >= 2)
paste("Number of Independants: ", nrow(Non2016Voters))
```

We have 659 people who claimed they Definitely did not vote or not compeltely sure they voted in 2016, to being certain they voted in 2018.

Now, we need to see the distribution of the general emotion rankings.

```{r, fig.width=8, fig.height=4}
hist(Non2016Voters$geangry, main = "Histogram of Anger", xlab = "Setimate rating")
hist(Non2016Voters$geafraid, main = "Histogram of Anger", xlab = "Setimate rating")
```


```{r}
Non2016Voters$angry <- ifelse(Non2016Voters$geangry >= 4, 1, 0)
Non2016Voters$afraid <- ifelse(Non2016Voters$geafraid >= 4, 1, 0)
```

```{r}
more_anger = sum( Non2016Voters$afraid < Non2016Voters$angry)
trials = sum( Non2016Voters$afraid < Non2016Voters$angry | Non2016Voters$afraid > Non2016Voters$angry)
binom.test(more_anger , trials)
```

No difference between fear or anger.
