---
title: "Lab 1: Comparing Means"
author: 'Srishti Mehra'
output:
  pdf_document: default
---

```{r, include=FALSE}
options(tinytex.verbose = TRUE)
```


```{r}
A = read.csv("anes_pilot_2018.csv")
A20 = read.csv("anes_pilot_2020ets_csv.csv")
```
    
# Research Questions

## Question 1: Do US voters have more respect for the police or for journalists?

### Introducing the topic

> Concept: To understand if US voters have different amount of respect police and jounralists. What we are trying to measure here, is a subjective variable.  

> Operationalization: From a list of data points collected in the ANES 2018 Pilot Study, two are people's ratings of the police and of journalists in variables 'ftpolice' and 'ftjournal' respectively. The questions to collect these are "How would you rate the police?" and "How would you rate journalists?". These are the closest data points available to our question since they describe to what the sample population feels about the police and journalists respectively. The answers to these questions are collected on a rating scale between 0 to 100.

> To get the subset of voters, we will use the voted2018 as descibed in the introduction.

> Concerns or Gaps:

> 1. The data collected is for rating police and journalists in general, not particularly based on respect. We have to be cautious while interpreting our conclusions since we are inferring respect from a total rating.

> 2. One person's interpretation of the rating scale can be very different from another's, so if one person could think of brilliant as 90, another could think of brilliant as 94. This must also be kept in mind as caution while reading the conclusion.

### Exploratory data analysis (EDA) of the relevant variables

> Variables: ftpolice, ftjournal

> Types: Both ftpolice and ftjournal are Ordinal type variables. Since the variables are measured on a rating scale, they have defined categories that have an order. So can apply operators like <, >, as well as =. We cannot however, apply any arithmetic operaters (like +, -, /) to such variables.
         
> Number of Entries 

```{r}
library(tidyverse)
usVoters <- A  %>% filter(A$voted2018  == 1)
paste("Number of Entries for ftpolice:", length(usVoters$ftpolice))
paste("Summary for ftpolice: ")
summary(usVoters$ftpolice)
paste("Number of Entries for ftjournal:", length(usVoters$ftjournal))
paste("Summary for ftjournal: ")
summary(usVoters$ftjournal)
```

> There are no Null Values for ftpolice and ftjournal

```{r}
paste("Number of Null Values for ftpolice:", length(which(is.na(usVoters$ftpolice))))
paste("Number of Null Values for ftjournal:", length(which(is.na(usVoters$ftjournal))))
```


> We will draw a histogram of both ftpolice and ftjournal to observe the distribution and any outliers

```{r}
hist(usVoters$ftpolice, breaks = 100,
     main = "Histogram of US voters' rating of the Police",
     xlab = "Rating of Police on a scale from 
     0(Very cold or unfavorable feeling) to 100 (Very warm or favorable feeling)")
```
> The distribution for ftpolice seems a little skewed towards the higher end, with upticks in the center.

```{r}
hist(usVoters$ftjournal, breaks = 100,
     main = "Histogram of US voters' rating of the Journalists",
     xlab = "Rating of Journalists on a scale from 
     0(Very cold or unfavorable feeling) to 100 (Very warm or favorable feeling)")

```

> The distribution for ftjournal is lesser skewed than that of ftpolice. 

> With the number of observations in both samples being well above the thumb rule (n=30), we can use the Central Limit Theorem to deduce that the distribution of the means approaches a normal distribution.

### Based on our EDA, selecting an appropriate hypothesis test

> 1. Since we are comparing two samples, it will be a two-sample test. 
> 2. Since the variables are of ordinal scale, we will use a non-parametric test.
> 3. Since the two variables we are looking at are answered by the same person, there is enough dependence that we can take advantage of if we do a paired test. 
> With all these above reasons, we narrow our choice of test down to Sign Test.

> Assumptions to be able to use the Sign Test are:
> 1. Variables represented are of ordnial Scale
> 2. Data is paired and drawn from i.i.d samples

> Since both the assumptions are satisfied, we will go ahead with using the sign test.

> The Null Hypothesis we will be testing is:
> US voters have equal respect for the police and journalists

> We will do a two-tailed test since we are only interested in the alternative hypothesis:
> US voters do not have equal respect for the police and journalists

### Conducting the chosen test

```{r}
more_police = sum( usVoters$ftjournal < usVoters$ftpolice)
trials = sum( usVoters$ftjournal < usVoters$ftpolice | usVoters$ftjournal > usVoters$ftpolice)
binom.test(more_police , trials)
```

> The p-value being well under 0.05 (statistical significance we decided to consider) indicates that we can reject the null hypothesis that US voters have equal respect for the police and journalists.

> Practical significance:
> The number of US voters in this sample who respect police more divided by the total number pairs of respecting either police or journalists more is 0.56. This is instead of the 0.5 we assumed for our null hypothesis, which would have reflected that US voters have equal respect for the police and journalists.




## Question 2: Are Republican voters older or younger than Democratic voters?

### Introducing the topic

> Concept: To measure the difference in ages of voters who identified as Republican or Democratic, we can use a metric variable because age is (just!) a number. 

> Operationalization: We have to find a variable that shows if the voter identifies as Republican or Democratic. The variable 'pid1d' captures the result of the question "Generally speaking, do you usually think of yourself as a Democrat, a Republican" in the survey matches closely to what we want and thus, we will use it. This variable takes one of the following six values:
> -7 No Answer
> -1 inapplicable, legitimate skip
> 1 Democrat
> 2 Republican
> 3 independent
> 4 something else

> There is another variable 'pid7x' which is a part of YouGov profile survey data that contains this affiliation. However, since these data were collected on previously-completed questionnaires, we identify the value in 'pid1d' being more recent and therefore will use that for the analysis our question. 

> For the age, we can use the 'birthyr' variable that is captured as a part of the survey, subtract that from the current year and get age. 

### Exploratory data analysis (EDA) of the relevant variables

> Variables: birthyr, pid1d

> Types: pid1d is a nominal level variable, however, we are only using that to separate our two samples that we are going to compare the ages for. Age is a ratio level variable.
         
> Number of Entries 

```{r}
currentYr <- lubridate::isoyear(Sys.Date())
usVoters$age <- currentYr-usVoters$birthyr
democrats <- usVoters  %>% filter(usVoters$pid1d==1)
republicans <- usVoters  %>% filter(usVoters$pid1d==2)

paste("Number of Entries for age of democrats:", length(democrats$age))
paste("Summary for age of democrats: ")
summary(democrats$age)
paste("Number of Entries for age of republicans:", length(republicans$age))
paste("Summary for age of republicans: ")
summary(republicans$age)
```

> There are no Null Values for ages of democrats or republicans

```{r}
paste("Number of Null Values for ages of democrats:", length(which(is.na(democrats$age))))
paste("Number of Null Values for ages of republicans:", length(which(is.na(democrats$age))))
```

> There are no duplicate rows in the dataset of US Voters as we had already established while answering research question 1

> We will draw a histogram of ages of democrats to observe the distribution and any outliers

```{r}
hist(democrats$age)
```

> To see the deviation from normality, we will plot a qqnorm plot

```{r}
qqnorm(democrats$age)
```

> We will observe the skewness of age of democrats

```{r}
skewness(democrats$age)
```

> The distribution for age of democrats is not far from normal, and with the number of observations being well above the thumb rule and the skewness being at at a low level (skewness is between -0.5 and 0.5), we can use the Central Limit Theorem to deduce that the distribution of the means approaches a normal distribution.

> We will draw a histogram of ages of republicans to observe the distribution and any outliers

```{r}
hist(republicans$age)
```

> To see the deviation from normality, we will plot a qqnorm plot

```{r}
qqnorm(republicans$age)
```

> We will observe the skewness of age of republicans

```{r}
skewness(republicans$age)
```

> The distribution for age of republicans is not far from normal either, and with the number of observations being well above the thumb rule and the skewness being at a low level (skewness is between -0.5 and 0.5), we can use the Central Limit Theorem to deduce that the distribution of the means approaches a normal distribution.

> As discussed earlier the sample is iid drawn. However, we must be cautious in this set with the age as the respondents of lower ages might be more motivated (because of the money they can earn through it) to respond than those of higher ages. It will apply to both the sets - democrats and republicans. 

### Based on our EDA, selecting an appropriate hypothesis test

>We will compare the means for the two samples. However, we should know that the mean alone cannot represent the enture distribution and answer our question. The conclusion thus, must be read with that in mind. 

> 1. Since we are comparing means of two samples, it will be a two-sample test. 
> 2. Since the variable we are comparing is a metric variable, parametric tests can be used.
> 3. Since the two variables we are looking at are answered by the different people and there is no dependency we can take advantage of, we will do a unpaired test. 
> With all these above reasons, we narrow our choice of test down to an unpaired t-test.

> Assumptions to be able to use the unpaired t-test are:
> 1. Metric variable
> 2. Data is paired and drawn from i.i.d samples
> 3. Distribution of the variable is (not too un-)normal

> Since all the assumptions are satisfied, we will go ahead with using the unpaired t-test.

> The Null Hypothesis we will be testing is:
> Republican voters are no more older or younger than Democratic voters

> We will do a two-tailed tests since we are only interested in the alternative hypothesis:
> Republican voters are either older or younger than Democratic voters

> We will consider the significance level of 0.05 since we are looking at a only a few hundred observations. Since the number of observations is not very large here, we will consider rejecting the null hypothesis if the p-value is <0.05.

### Conducting the chosen test

```{r}
t.test(democrats$age, republicans$age, paired=F)
```

> Since the p-value is less than 0.05, the test is statistically significant to reject the null hypothesis per the significance level we considered before doing the test.

> Practical significance:

```{r}
meanage_dem = mean(democrats$age)
sdage_dem = sd(democrats$age)
paste("Mean of Age of Democrats: ", meanage_dem, " Sampling Standard Deviation in Age Democates: ", sdage_dem)

meanage_rep = mean(republicans$age)
sdage_rep = sd(republicans$age)
paste("Mean of Age of Republicans: ", meanage_rep, " Sampling Standard Deviation in Age Republicans: ", sdage_rep)
```

> The difference in means of ages of Democrats and Republicans is 3 with a small difference in sampling standard deviations too (0.3). However, from the graph below of the sampling distribution of the ages of democrats and republicans, we see a substancial area overlapping, so there is practical significance to reject the null hypothesis is not very high.  

```{r}
library(ggpubr)

dems_reps <- usVoters %>%   filter(usVoters$pid1d==2 | usVoters$pid1d==3)
dems_reps$party <- ifelse(dems_reps$pid1d==2,'democrat','republican')
ggdensity(dems_reps, x = "age", fill = "party") +
  scale_fill_manual(values=c("#222a84","#cf1c10"))

```

## Question 3: Do a majority of independent voters believe that the federal investigations of Russian election interference are baseless?

### Introducing the topic

> Concept: To measure voters' belief about whether federal investigations of Russian election interference are baseless or not, we would use a nominal variable with values indicating true or false. We will measure this for voters who identify their party inclination as independent.

> Operationalization: We will use the 'pid1d' variable that captures the answer to "Generally speaking, do you usually think of yourself as a Democrat, a Republican?" to see if a voter identifies as independent. As mentioned in the earlier question's analysis, this variable takes one of six values. Here, we will be interested in only one of them: 3 which indicates independent.

> For the belief about whether federal investigations of Russian election interference are baseless or not, there are two closely associated variables we can use:
> 1. 'russia16' which captures the answer to question "Do you think the Russian government probably interfered in the 2016 presidential?" 
> 2. 'muellerinv' which captures the answer to question "Do you approve, disapprove, or neither approve nor disapprove of Robert Mueller’ [ATTENTION HERE!]"

> The 'russia16' variable takes one of three values:
> -7 No Answer
> 1 Russia probably interfered
> 2 This probably did not happen

>While this variable does not capture the belief of the federal investigations being baseless or not, it is likely that if a voter things there was inteference (indicated by the variable 'russia16'), they will not find the investigations baseless and vice versa. 

> The 'muellerinv' variable takes one of the following eight values:
> -7 No Answer
> 1 Approve extremely strongly
> 2 Approve moderately strongly
> 3 Approve slightly
> 4 Neither approve nor disapprove
> 5 Disapprove slightly
> 6 Disapprove moderately strongly
> 7 Disapprove extremely strongly

> Concerns or Gaps:

> 1. The data collected is for two questions that are not exactly what we are trying to analyze. They are the ones we think are closely related and informative about the variable we are trying to analyze.  
> 2. Again, one person's interpretation of the Likert scale can be very different from another's. One person's 'Disapprove slightly' could be the same as another's 'Disapprove moderately strongly'. 

### Exploratory data analysis (EDA) of the relevant variables

> Variables: russia16, pid1d

> Types: pid1d is a nominal level variable, however, we are only using that to get our sample of independent voters. russia16 is also a nominal level variable. muellerinv is an ordinal variable since the values can be ordered from the level of approval to the level of disapproval.
         
> Number of Entries 

```{r}
independents <- usVoters  %>% filter(usVoters$pid1d==3)

paste("Number of Entries for the variable russia16 of independents:", length(independents$russia16))

paste("Number of Entries for the variable muellerinv of independents:", length(independents$muellerinv))
```

> This makes sense because the possible values the value can take are only 1 or 2. From just looking at the mean, it seems that there are more 1s than 2s. 

> There are no Null Values for the variable russia16 or muellerinv for independents

```{r}
paste("Number of Null Values for the variable russia16 of independents:", length(which(is.na(independents$russia16))))
paste("Number of Null Values for the variable muellerinv of independents:", length(which(is.na(independents$muellerinv))))
```

> There are no duplicate rows in the dataset of US Voters as we had already established while answering research question 1

> We will not evaluate the histogram or skewness of the russia16 variable for independent voters because it is a nominal variable and the histogram or skewness will not inform us much.

> We will plot the histogram of muellerinv for independent voters to see it's distribution 

```{r}
hist(independents$muellerinv)
```

> We will observe the skewness of muellerinv of independent voters

```{r}
skewness(independents$muellerinv)
```

> The distribution for muellerinv for independent voters seems symmetric about 4 and the skewness level is at a low level (skewness is between -0.5 and 0.5,).

> As discussed earlier the sample is iid drawn.  

> Since mullerinv seems a closer informative of our question and it has an ordinal level, we will use that for our analysis. 

### Based on our EDA, selecting an appropriate hypothesis test
cw our choice of test down to the Wilcoxon rank-sum test.

> Assumptions to be able to use the Wilcoxon rank-sum test are:
> 1. Ordinal variable
> 2. Data is paired and drawn from i.i.d samples

> Since all the assumptions are satisfied, we will go ahead with using the Wilcoxon rank-sum test.

> The Null Hypothesis we will be testing is:
> Independent voters are no more likely to believe that the federal investigations of Russian election interference are baselesss than they are to believe that the federal investigations of Russian election interference are not baseless

> We will do a two-tailed tests even though we are interested in whether a majority of independent voters believe that the federal investigations of Russian election interference are baseless because:
> 1. We must choose what kind of test we are going to use before looking at the data. The use of one-tailed tests (which are lesser common) needs to be justified to large extents. We have to be cautious of the fact that it might not be widely accepted, for example, the audience might not believe that we started with a one-tail before looking at the data and think that we changed my test after running it once with two-tail and not getting a statistically significant result. Or the audience in general might not share the same opinion on my justifications for the one-tailed test.
> 2. With the one-tailed test, it is easier to reject the null hypothesis given all of the rejection region is on one side -> it is therefore associated with more skepticism than a one-tailed test. 

> We will consider the significance level of 0.05 since we are looking at a only a few hundred observations. Since the number of observations is not very large here, we will consider rejecting the null hypothesis if the p-value is <0.05.

### Conducting the chosen test

```{r}
wilcox.test(independents$muellerinv)
```

> Since the p-value is less than 0.05, the test is statistically significant to reject the null hypothesis per the significance level we considered before doing the test.

> Practical significance:

```{r}
num_ind = length(independents$muellerinv)
paste("Number of independent voters: ",num_ind)
approved <- independents %>% filter(independents$muellerinv<4 & independents$muellerinv!=-7)
neutral <- independents %>% filter(independents$muellerinv==4)
disapproved <- independents %>% filter(independents$muellerinv>4)

paste("Ratio of independent voters in the sample who approve of the Mueller Investigation: ",length(approved$muellerinv)/num_ind)
paste("Number of independent voters in the sample who were nuetral about of the Mueller Investigation: ",length(neutral$muellerinv)/num_ind)
paste("Number of independent voters in the sample who disapprove of the Mueller Investigation: ",length(disapproved$muellerinv)/num_ind)
```

> This shows there is a higher ratio of independent voters who approve of the Mueller Investigation. We can deduce that they must think the federal investigations of Russian election interference were not baseless. Therefore, there is practical significanct effect we see as well. 

## Question 4: Was anger or fear more effective at driving increases in voter turnout from 2016 to 2018?

### Introducing the topic

> Concept: To measure the increase in turnout from 2016 to 2018, we the population of people who did not turnout in 2016 but did in 2018 since they will make the increase. We can also consider those who got eligible for voting in 2018 by their birth year. 
To gather which of the emotions, anger or fear, were driving factors for the increased turnout, we would have questions asking the respondents whether any or both of those feelings drove them to turnout to vote in 2018 when they did not in 2016. 

> Operationalization: Since we are concerned with the increase driven by anger or fear, we can safely discount the respondents who became eligible to vote for the 2018 election and were not eligble for the 2016 election. 
> The variables 'turnout16' and 'turnout18' which capture answers to "In 2016, the major candidates for president were Donald Trump for the Republicans and Hillary Clinton for the Democrats. In that election, did you definitely vote, definitely not vote, or are you not completely sure whether you voted?" and "In the election held on November 6, did you definitely vote in person on election day, vote in person before Nov 6, vote by mail, did you definitely not vote, or are you not completely sure whether you voted in that election?" respectively are what we can use to build the segment for increase as discussed in the concept.
> We can also include those respondents who mentioned they probably did not vote in 2016 reflected the variable 'turnout16b' which captures the answer to "Do you think you probably voted or probably did not vote?" And for the 2018 election, we can include those you mentioned probably did vote in 2018 reflected in the variable 'turnout18ns' which captures the answer to "If you had to guess, would you say that you probably did vote in the election held on
November 6, or probably did not vote in that election?"

> There are two variables captured in the ANES 2018 Pilot Study that represent "Generally speaking, how do you feel about the way things are going in the country these days?" with respect to anger and fear. 'geangry' captures "How angry do you feel?" and 'geafraid' captures "How afraid do you feel?". The variables take one of the following six values depending on how intensely they feel the emotion being represented:
> -7 No Answer
> 1 Not at all
> 2 A little
> 3 Somewhat
> 4 Very
> 5 Extremely

>Since we are interested in which of the two emotions were more effective in driving turnout, we will select the sample that provided an answer for both the variables. 

> Concerns or Gaps:

> 1. The data collected is for the feeling of anger and fear in general, not particularly whether it is effective in driving turnout. The question was not whether anger or fear drove the candidate to turnout in 2018 when they did not in 2016. 
> 2. One person's interpretation of the Likert scale can be very different from another's, so if one person could represent their immense fear as 'very', represented by 4 on the scale, another person could represent the same as 'extremely', represented by 5 on the scale. Similarly for anger as well. 

### Exploratory data analysis (EDA) of the relevant variables

> Variables: turnout16, turnout16b, turnout18, turnout18ns, geangry, geafraid

> Types: turnout16, turnout16b, turnout18, turnout18ns are all nominal variables, but we are only using that to get our sample. geangry and geafraid are ordinal variables since can be ordered from the least strong feeling of the emotion to most strong. 

> Number of Entries 

```{r}
increased_voters <- A  %>% filter((A$turnout18<4 | (A$turnout18==4 & A$turnout18ns==1)) & (A$turnout16==2 | (A$turnout16==2 & A$turnout16b==2)) & (A$geangry!=-7 & A$geafraid!=-7))

paste("Number of Entries for the variable geangry of increased_voters:", length(increased_voters$geangry))
paste("Summary for geangry: ")
summary(increased_voters$geangry)

paste("Number of Entries for the variable geafraid of increased_voters:", length(increased_voters$geafraid))
paste("Summary for geafraid: ")
summary(increased_voters$geafraid)
```

> There are no Null Values for geangry and geafraid

```{r}
paste("Number of Null Values for geangry:", length(which(is.na(increased_voters$geangry))))
paste("Number of Null Values for geafraid:", length(which(is.na(increased_voters$geafraid))))
```

> There are no duplicate rows in the dataset of US Voters as we had already established while answering research question 1

> Will draw a histogram of both geangry to observe the distribution and any outliers

```{r}
hist(increased_voters$geangry)
```

> We will now observe the skewness of geangry variable

```{r} 
skewness(increased_voters$geangry)
```

> We see a symmetry about the median (3) in the distribution for geangry and the skewness level is also low (skewness is between -1 and -0.5 or between 0.5 and 1).

> Will draw a histogram of both geafraid to observe the distribution and any outliers

```{r}
hist(increased_voters$geafraid)
```

> We will now observe the skewness of geafraid variable

```{r} 
skewness(increased_voters$geafraid)
```

> Again, we see a symmetry about the median (3) in the distribution for geafraid and the skewness level is also low (skewness is between -1 and -0.5 or between 0.5 and 1).

> As discussed earlier the sample is iid drawn. 

### Based on our EDA, selecting an appropriate hypothesis test

> 1. Since we are comparing two samples, it will be a two-sample test. 
> 2. Since the variables are of ordinal scale, parametric tests cannot be used.
> 3. Since the two variables we are looking at are answered by the same person, there is enough dependence that we can take advantage of if we do a paired test. 
> With all these above reasons, we narrow our choice of test down to Sign Test.

> Assumptions to be able to use the Sign Test are:
> 1. Variables represented are of ordnial Scale
> 2. Data is paired and drawn from i.i.d samples

> Since both the assumptions are satisfied, we will go ahead with using the sign test.

> The Null Hypothesis we will be testing is:
> Anger is no more or less effective than fear at driving increases in voter turnout from 2016 to 2018

> We will do a two-tailed test since we are only interested in the alternative hypothesis:
> There is a difference in the effectiveness of anger and fear at driving increases in voter turnout from 2016 to 2018

> We will consider the significance level of 0.05 since we are looking at a only a few hundred observations. Since the number of observations is not very large here, we will consider rejecting the null hypothesis if the p-value is <0.05.

### Conducting the chosen test

```{r}
more_anger = sum( increased_voters$geafraid < increased_voters$geangry)
trials = sum( increased_voters$geafraid < increased_voters$geangry | increased_voters$geafraid > increased_voters$geangry)
binom.test(more_anger , trials)
```

> The p-value is greater than 0.05 indicating that we cannot reject the null hypothesis that Anger is no more or less effective than fear at driving increases in voter turnout from 2016 to 2018.

> Practical significance:
> We see that the probability of increased voters having anger be more represented than fear comes up as 0.49 from the Sign Test. This also indicates that the practical significance of the effect is very low.


## Question 5: Select a fifth question that you believe is important for understanding the behavior of voters

### Introducing our Research Question

> There are a variety of issues people care about and consider significantly while voting and while choosing a candidate to vote for. Income inequality is one such issue. Income inequality has fluctuated over the years, both increasing and decreasing in portions of the last 20 years. 

> The question we are interested in is how Hillary voters view the change in income inequality over the last 20 years versus how Trump voters view the change in income inequality over the last 20 years.

> It has been argued that the extent and nature of participation in choosing leaders is closely associated with the distribution of resources in society (https://nathanjkelly.utk.edu/wp-content/uploads/2017/10/Franko-et-al-2016.pdf). It is plausible that voters' perception of change in income inequality could affect their voting behavior. Therefore, we feel this is an important question in understanding voting behavior.

> We will evaluate this by building the appropriate hypothesis and using appropriate statistical tests under the assumptions that make them valid. 

> Concept: To evaluate if Hillary voters look at the change in income inequality over the last 20 years differently from the way Trump voters look at the same. We are interested in seeing what people's perception of change in income inequality is in 2018 based on what candidate they preferred in 2016.  

> Operationlization: The variable 'richpoor' captures the answer to the question "Do you think the difference in incomes between rich people and poor people in the United
> States today is larger, smaller, or the same as it was 20 years ago?". This question was presented to all the participants of the survey. It can take one of the following seven values (along with the representing choices that the participants given to choose from):
1 - A lot larger
2 - A moderate amount larger
3 - A little larger 
4 - The same
5 - A little smaller 
6 - A moderate amount smaller
7 - A lot smaller
We will use this variable to understand how the voters viewed the change in income inequality in the last 20 years.

> The variable 'vote16' captures the answer to the questions "In the 2016 presidential election, who did you vote for? Donald Trump, Hillary Clinton, or
someone else?". It takes one of the following three values (along with the representing choices that the participants given to choose from):
1 - Donald Trump
2 - Hillary Clinton
3 - Someone else
> We will use this to distinguish between Hillary voters and Trump voters. 


### Exploratory data analysis (EDA) of the relevant variables

> Variables: richpoor, vote16

> Types: richpoor is an ordinal variable since it has categories that have an order. Operataions like >,<,= are valid for the 'richpoor' variable. vote16 is a nominal variable since it has categories that do not have an order. Operataions like >,<,= do not make sense for the vote16 variable. We are only using the vote16 variable to separate out the two samples we are interested in comparing. 
         
> Number of Entries 

```{r}
hillary_voters <- A %>% filter(A$vote16==2)
trump_voters <- A %>% filter(A$vote16==1)

paste("Number of voters who voted for Hillary Clinton in 2016:", length(hillary_voters$vote16))

paste("Number of voters who voted for Donald Trump in 2016:", length(trump_voters$vote16))
```

> There are no Null Values for the variable richpoor for Hillary or Trump voters

```{r}
paste("Number of Null Values for richpoor for Hillary voters:", length(which(is.na(hillary_voters$richpoor))))
paste("Number of Null Values for richpoor for Trump voters:", length(which(is.na(trump_voters$richpoor))))
```

> Drawing out the histograms for the richpoor variable for both Hillary and Trump voters to observe sampling distribution

```{r}
hist(hillary_voters$richpoor)
hist(hillary_voters$richpoor, breaks = 7,
     main = "How Hillary voters viewed change in income inequality",
     xlab = "Change in income inequality (ranging from 1-A lot larger to 7-A lot smaller)")
```

```{r}
hist(trump_voters$richpoor)
hist(trump_voters$richpoor, breaks = 7,
     main = "How Trump voters viewed change in income inequality",
     xlab = "Change in income inequality (ranging from 1-A lot larger to 7-A lot smaller)")
```

### Based on our EDA, selecting an appropriate hypothesis test

> 1. Since we are comparing two samples, it will be a two-sample test. 
> 2. Since the variables are of ordinal scale, parametric tests cannot be used.
> 3. There is no dependence to rely on and therefore we will use an unpaired test. 
> The distributions from the histograms look slightly skewed towards the left however, since we are intending to use the Wilcoxon Rank-Sum test, the skewness does not affect our choice of test.

> With all these above reasons, we narrow our choice of test down to Wicoxon rank-sum test.

> Assumptions to be able to use the Sign Test are:
> 1. Variables represented are of ordnial Scale
> 2. Data is paired and drawn from i.i.d samples

> Since both the assumptions are satisfied, we will go ahead with using the Wicoxon rank-sum test.

> The Null Hypothesis we will be testing is:
> Hillary voters do not view change in income inequality over the last 20 years any differently than Trump voters do. 

> We will do a two-tailed test since we are only interested in the alternative hypothesis:
> There is a difference in the way Hillary voters and Trump voters view change in income inequality over the last 20 years.

### Conducting the chosen test

```{r}
wilcox.test(trump_voters$richpoor,hillary_voters$richpoor)
```

> The p-value is less than 0.05 indicating that this test is statistically significant to reject the null hypothesis that Hillary voters do not view change in income inequality over the last 20 years any differently than Trump voters do.

### Conclusion

> From this test we see that Hillary voters view the change income inequality over the past 20 years differently from how Trump voters view this change.

> We have noticed that the histograms we saw for the sample distributions reflected differences in how Hillary voters perceived change in income inequality over the past 20 years versus how Trump voters did. More Hillary voters thought of the inequality to have become a lot larger whereas Trump voters had opinions more spread on the scale.

> From the wilcox.test documentation, "R's value can also be computed as the number of all pairs (trump_voters$richpoor, hillary_voters$richpoor) for which hillary_voters$richpoor is not greater than trump_voters$richpoor, the most common definition of the Mann-Whitney test."
>
> The total number of pairs is,

```{r}
(sum(A$vote16 == 1 & !is.na(A$richpoor)) * sum(A$vote16 == 2 & !is.na(A$richpoor)))
```

> 536432 (value of W obtained from our test) out 691460 pairs showed that there was a shift in the percieved rating for change in income inequality between Trump and Hillary voters. The percentage that gets us is 

```{r}
paste(536432*100/691460, "%")
```

>77% is a percent we feel comfortable to use and say that this shows that the test is practically significant along with being statistically significant. 

> Common wisdom is that Democrat voters are more concerned about issues of inequality as compared to Republican voters and test results are consistent with this view.