---
title: "Lab 1: Comparing Means"
author: 'Srishti Mehra'
output:
  pdf_document: default
---

```{r, include=FALSE}
options(tinytex.verbose = TRUE)
```


```{r}
A = read.csv("anes_pilot_2018.csv")
```
    
# Research Questions

## Question 1: Do US voters have more respect for the police or for journalists?

### Introducing the topic

> Concept: To measure the amount of respect US voters have for police and jounralists, since it is subjective, we cannot use a metric variable. We have to use a scale like the Likert scale to possibly capture this data and therefore it can be best defined with an ordinal variable. 


> Operationalization: From a list of data points collected in the ANES 2018 Pilot Study, two are people's ratings of the police and of journalists in variables 'ftpolice' and 'ftjournal' respectively. The questions to collect these are "How would you rate the police?" and "How would you rate journalists?". These are the closest data points available to our question since they are the only ones related to what the sample population feels about the police and journalists respectively. The answers to these questions are collected on a Likert scale.
>We are also going to take a subset of the total sample available from the ANES 2018 Pilot study of those people who have reported that they voted. To do so we will consider the variable 'turnout18' which contains the answer to the question "In the election held on November 6, did you definitely vote in person on election?". The variable holds one of the following five values:

> 1 - Definitely voted in person on Nov 6
> 2 - Definitely voted in person, before Nov 6
> 3 - Definitely voted by mail
> 4 - Definitely did not vote
> 5 - Not completely sure

> Another variable that could inform a portion of the population who have the 'turnout18' variable value as 5 who could have possibly voted can be covered from the variable 'turnout18ns'. This variable captures answer to the question "If you had to guess, would you say that you probably did vote in the election?". This variable takes one of the following three values: 

> 1 - inapplicable, legitimate skip
> 2 - Probably did vote
> 3 - Probably did not vote

> For our analysis, we will only consider the population that have value 1, 2, or 3 for the variable 'turnout18' to take the conserative approach in considering US voters.


> Concerns or Gaps:

> 1. The data collected is for rating police and journalists in general, not particularly based on respect. Inferring respect from a total rating will not give us accurate conclusions to draw.
> 2. One person's interpretation of the Likert scale can be very different from another's, so if one person could think of brilliant as 90, another could think of brilliant as 94. This must be kept in mind as caution while reading the conclusion.

### Exploratory data analysis (EDA) of the relevant variables

> Variables: ftpolice, ftjournal

> Types: Both ftpolice and ftjournal are Ordinal type variables. Since the variables are measured on a Likert scale, they have defined categories that have an order. So can apply operators like <, >, as well as =. However, t
         
> Number of Entries 

```{r}
library(tidyverse)
usVoters <- A  %>% filter(A$turnout18 > 0 & A$turnout18 < 4)
paste("Number of Entries for ftpolice:", length(usVoters$ftpolice))
paste("Summary for ftpolice: ")
summary(usVoters$ftpolice)
paste("Number of Entries for ftjournal:", length(usVoters$ftjournal))
paste("Summary for ftjournal: ")
summary(usVoters$ftjournal)
```

> There are no Null Values for ftpolice and ftjournal

```{r}
paste("Number of Null Values for ftpolice:", length(which(is.na(usVoters$ftpolice))))
paste("Number of Null Values for ftjournal:", length(which(is.na(usVoters$ftjournal))))
```

> There are no duplicate rows in the dataset of US Voters

```{r}
paste("Number of Duplicate Values for the dataset usVoters:", length(which(duplicated(usVoters))))
```

> Will draw a histogram of both ftpolice and ftjournal to observe the distribution and any outliers

```{r}
hist(usVoters$ftpolice)
```

> We will now plot and observe the skewness of ftpolice variable

```{r}
library(e1071)  
skewness(usVoters$ftpolice)
```

> The distribution for ftpolice is not normal, however with the number of observations being well above the thumb rule and the skewness is at a moderate level (skewness is between -1 and -0.5 or between 0.5 and 1), we can use the Central Limit Theorem to deduce that the distribution of the means approaches a normal distribution.

```{r}
hist(usVoters$ftjournal)
```

> We will now plot and observe the skewness of ftpolice variable

```{r}
library(e1071)  
skewness(usVoters$ftjournal)
```

> The distribution for ftjournal is not normal, however with the number of observations being well above the thumb rule and the skewness is at a low level (skewness is between -0.5 and 0.5,), we can use the Central Limit Theorem to deduce that the distribution of the means approaches a normal distribution.

> We will now observe the variable to see if it is identically and independently distributed (iid)

>  As a modern survey, the ANES 2018 Pilot Study we can expect the selection to be from iid drawn.  You might worry that some people are more likely to respond than others (some more motivated for the money since this is a survey they get paid for).  But this isn't a problem for iid, it just means that the population of respondents is not the same as the population representing all Americans.  It's a problem for generalizability. Since there is only a finite number of Americans, it is impossible to have perfect independence.  Once a person is drawn, they cannot be drawn again, so the distribution for the next draw has changed.  However, the change is likely to be very small, so we can safely ignore these finite-sample effects.  On the whole, there are issues to worry about, but modern surveys like the ANES 2018 Pilot Study do a pretty good job of making the iid assumption valid. Codebook and User’s Guide to the ANES 2018 Pilot Study mentiones that the survey participation is independent of variables measured in the survey. 

### Based on our EDA, selecting an appropriate hypothesis test

> 1. Since we are comparing means of two samples, it will be a two-sample test. 
> 2. Since the variables are of ordinal scale, parametric tests cannot be used.
> 3. Since the two variables we are looking at are answered by the same person, there is enough dependence that we can take advantage of if we do a paired test. 
> With all these above reasons, we narrow our choice of test down to Sign Test.

> Assumptions to be able to use the Sign Test are:
> 1. Variables represented are of ordnial Scale
> 2. Data is paired and drawn from i.i.d samples

> Since both the assumptions are satisfied, we will not go ahead with using the sign test.

> The Null Hypothesis we will be testing is:
> US voters have equal respect for the police and journalists

> We will do a two-tailed tests since we are only interested in the alternative hypothesis:
> US voters do not have equal respect for the police and journalists

### Conducting the chosen test

```{r}
more_police = sum( usVoters$ftjournal < usVoters$ftpolice)
trials = sum( usVoters$ftjournal < usVoters$ftpolice | usVoters$ftjournal > usVoters$ftpolice)
binom.test(more_police , trials)
```

> The p-value being well under 0.01 (statistical significance levels generally acceptable) indicating that we can reject the null hypothesis that US voters have equal respect for the police and journalists.

>To now see the statistical significance:
> We see that the probability of US voters have respecting police more comes up as 0.56 from the Sign Test. In our hypothesis we assumed it would be 0.5 reflecting that US voters have equal respect for the police and journalists. Thus, we see that the effect is not a very large one even though the p-value was very low to reflect a statistically significant result.


## Question 2: Are Republican voters older or younger than Democratic voters?

### Introducing the topic

> Concept: To measure the difference in ages of voters who identified as Republican or Democratic, we can use a metric variable because age is (just!) a number. 

> Operationalization: We have to find a variable that shows if the voter identifies as Republican or Democratic. The variable 'pid1d' captures the result of the question "Generally speaking, do you usually think of yourself as a Democrat, a Republican" in the survey matches closely to what we want and thus, we will use it. This variable takes one of the following six values:
> -7 No Answer
> -1 inapplicable, legitimate skip
> 1 Democrat
> 2 Republican
> 3 independent
> 4 something else

> There is another variable 'pid7x' which is a part of YouGov profile survey data that contains this affiliation. However, since these data were collected on previously-completed questionnaires, we identify the value in 'pid1d' being more recent and therefore will use that for the analysis our question. 

> For the age, we can use the 'birthyr' variablew that is captured as a part of the survey, subtract that from the current year and get age. 

### Exploratory data analysis (EDA) of the relevant variables

> Variables: birthyr, pid1d

> Types: pid1d is a nominal level variable, however, we are only using that to separate our two samples that we are going to compare the ages for. Age is a ratio level variable.
         
> Number of Entries 

```{r}
library(tidyverse)
currentYr <- lubridate::isoyear(Sys.Date())
usVoters$age <- currentYr-usVoters$birthyr
democrats <- usVoters  %>% filter(usVoters$pid1d==2)
republicans <- usVoters  %>% filter(usVoters$pid1d==3)

paste("Number of Entries for age of democrats:", length(democrats$age))
paste("Summary for age of democrats: ")
summary(democrats$age)
paste("Number of Entries for age of republicans:", length(republicans$age))
paste("Summary for age of republicans: ")
summary(republicans$age)
```

> There are no Null Values for ages of democrats or republicans

```{r}
paste("Number of Null Values for ages of democrats:", length(which(is.na(democrats$age))))
paste("Number of Null Values for ages of republicans:", length(which(is.na(democrats$age))))
```

> There are no duplicate rows in the dataset of US Voters as we had already established while answering research question 1

> We will draw a histogram of ages of democrats to observe the distribution and any outliers

```{r}
hist(democrats$age)
```

> To see the deviation from normality, we will plot a qqnorm plot

```{r}
qqnorm(democrats$age)
```

> We will now plot and observe the skewness of age of democrats

```{r}
library(e1071)  
skewness(democrats$age)
```

> The distribution for age of democrats is not far from normal, and with the number of observations being well above the thumb rule and the skewness is at a low level (skewness is between -0.5 and 0.5), we can use the Central Limit Theorem to deduce that the distribution of the means approaches a normal distribution.

> We will draw a histogram of ages of republicans to observe the distribution and any outliers

```{r}
hist(republicans$age)
```

> To see the deviation from normality, we will plot a qqnorm plot

```{r}
qqnorm(republicans$age)
```

> We will now plot and observe the skewness of age of republicans

```{r}
library(e1071)  
skewness(republicans$age)
```

> The distribution for age of republicans is not far from normal, and with the number of observations being well above the thumb rule and the skewness is at a low level (skewness is between -0.5 and 0.5), we can use the Central Limit Theorem to deduce that the distribution of the means approaches a normal distribution.

> As discussed earlier the sample is iid drawn. However, we must be cautious in this set with the age as the respondents of lower ages might be more motivated (because of the money they can earn through it) to respond than those of higher ages. It will apply to both the sets - democrats and republicans. 

### Based on our EDA, selecting an appropriate hypothesis test

> 1. Since we are comparing means of two samples, it will be a two-sample test. 
> 2. Since the variable we are comparing is a metric variable, parametric tests can be used.
> 3. Since the two variables we are looking at are answered by the different people and there is no dependency we can take advantage of, we will do a unpaired test. 
> With all these above reasons, we narrow our choice of test down to an unpaired t-test.

> Assumptions to be able to use the unpaired t-test are:
> 1. Metric variable
> 2. Data is paired and drawn from i.i.d samples
> 3. Distribution of the variable is (not too un-)normal

> Since all the assumptions are satisfied, we will not go ahead with using the signunpaired t-test.

> The Null Hypothesis we will be testing is:
> Republican voters are no more older or younger than Democratic voters

> We will do a two-tailed tests since we are only interested in the alternative hypothesis:
> Republican voters are either older or younger than Democratic voters

### Conducting the chosen test

```{r}
t.test(democrats$age, republicans$age, paired=F)
```

> Since the p-value is greater than 0.05 (therefore, also greater than 0.01), the test is not statistically significant to reject the null hypothesis

> Practical Significance:

```{r}
meanage_dem = mean(democrats$age)
sdage_dem = sd(democrats$age)
paste("Mean of Age of Democrats: ", meanage_dem, " Sampling Standard Deviation in Age Democates: ", sdage_dem)

meanage_rep = mean(republicans$age)
sdage_rep = sd(republicans$age)
paste("Mean of Age of Republicans: ", meanage_rep, " Sampling Standard Deviation in Age Republicans: ", sdage_rep)
```

> The difference in means of ages of Democrats and Republicans is only 0.82 with a small difference in sampling standard deviations too (0.3). And, we can from the graph below, a substancial area overlapping, so there is no practical significance to reject the null hypothesis either. 

```{r}
library(ggpubr)

dems_reps <- usVoters %>%   filter(usVoters$pid1d==2 | usVoters$pid1d==3)
dems_reps$party <- ifelse(dems_reps$pid1d==2,'democrat','republican')
ggdensity(dems_reps, x = "age", fill = "party", palette = "jco")

```


## Question 3: Do a majority of independent voters believe that the federal investigations of Russian election interference are baseless?

### Introduce your topic briefly.  (5 points)
Explain how your variables are operationalized.  Comment on any gaps that you can identify between your operational definitions and the concepts you are trying to study.

### Perform an exploratory data analysis (EDA) of the relevant variables. (5 points)
This should include a treatment of non-response and other special codes, basic sanity checks, and a justification for any values that are removed.  Use visual tools to assess the relationship among your variables and comment on any features you find.

### Based on your EDA, select an appropriate hypothesis test.  (5 points)
Explain why your test is the most appropriate choice.  List and evaluate all assumptions for your test.

### Conduct your test. (5 points)
Explain (1) the statistical significance of your result, and (2) the practical significance of your result.  Make sure you relate your findings to the original research question.


## Question 4: Was anger or fear more effective at driving increases in voter turnout from 2016 to 2018?

### Introduce your topic briefly.  (5 points)
Explain how your variables are operationalized.  Comment on any gaps that you can identify between your operational definitions and the concepts you are trying to study.

### Perform an exploratory data analysis (EDA) of the relevant variables. (5 points)
This should include a treatment of non-response and other special codes, basic sanity checks, and a justification for any values that are removed.  Use visual tools to assess the relationship among your variables and comment on any features you find.

### Based on your EDA, select an appropriate hypothesis test.  (5 points)
Explain why your test is the most appropriate choice.  List and evaluate all assumptions for your test.

### Conduct your test. (5 points)
Explain (1) the statistical significance of your result, and (2) the practical significance of your result.  Make sure you relate your findings to the original research question.


## Question 5: Select a fifth question that you believe is important for understanding the behavior of voters

### Clearly argue for the relevance of this question.  (10 points)
In words, clearly state your research question and argue why it is important for understanding the recent voting behavior. Explain it as if you were presenting to an audience that includes technical and non technical members.

Explain how your variables are operationalized.  Comment on any gaps that you can identify between your operational definitions and the concepts you are trying to study.

### Perform EDA and select your hypothesis test (5 points)

Perform an exploratory data analysis (EDA) of the relevant variables.

This should include a treatment of non-response and other special codes, basic sanity checks, and a justification for any values that are removed.  Use visual tools to assess the relationship among your variables and comment on any features you find.

Based on your EDA, select an appropriate hypothesis test.
Explain why your test is the most appropriate choice.  List and evaluate all assumptions for your test.

### Conduct your test. (2 points)
Explain (1) the statistical significance of your result, and (2) the practical significance of your result.

### Conclusion (3 points)
Clearly state the conclusion of your hypothesis test and how it relates to your research question.

Finally, briefly present your conclusion in words as if you were presenting to an audience that includes technical and non technical members.



